---
title: "Custom Label Fonts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom Label Fonts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

## Overview 
Obviously, everyone wants to be able to customize their label templates! 
Why go through the bother of installing and running `BarnebyLives` if you cannot do that!  

Note that the labels are made using LaTeX, an awesome guide for LaTeX (relevant to the level you will be using it at) is [Overleaf.org](https://www.overleaf.com/learn/latex/Learn_LaTeX_in_30_minutes), and check the links in the side bar too! 

In this vignette we will replace the default LaTeX fonts with add-on fonts using the `fontspec` package. 
Also note that previously you *may* have been able to scrape by using any old LaTeX installation, at this point you will definitely need to make sure you have either XelaTex for lualatext installed. 
'tinytex' which is installable directly from R should accomplish this on all operating systems. 

This vignette assumes you were able to get through `Custom Label Templates`. 

```{r tinytex for all OS, eval = F}
install.packages("tinytex")
tinytex::install_tinytex()
```

You can check that you have successfully installed this with
```{r}
tinytex::is_tinytex()
```

If you are on linux for the sake of compatibility for this vignette, please install a ton of Windows fonts. 
It seems most of these are on Mac - but if you start hitting errors, maybe give this a go. 

You can do this with the code below, which will give part of your soul to Bill Gates of yesteryear. 

```{sh, eval=F}
echo "ttf-mscorefonts-installer msttcorefonts/accepted-mscorefonts-eula select true" | sudo debconf-set-selections
sudo apt install -y ttf-mscorefonts-installer
```

or you can 
```{sh, eval=F}
sudo apt install ttf-mscorefonts-installer
```

and then click `tab` and `enter` on your keyboard when the debconf pops up to accept that you are giving Bill Gates of yesteryear your soul. 

```{sh}
fc-list | grep "TImes New Roman"
```

and doubly verify the Tex install can access them
```{sh}
sudo fc-cache -fv
```

You may get messages about loops, these are fine as long as the last line reads something like 

`fc-cache: succeeded`

## Prepare workspace

```{r load libraries and fine labels, eval =F}
library(BarnebyLives)
library(tidyverse)

local <- file.path('~', 'Documents', 'assoRted', 'Barneby_Lives_dev', 'LabelStyles')

l.nevadense <- collection_examples |>
  filter(Collection_number == 2823)

write.csv(l.nevadense, file.path(local, 'SoS-ExampleCollection.csv'))
```

All of the labels can be copied from their original locations using the following code.  

```{r copy a template to a local directory, eval = F}
p2lib <- file.path(
  system.file(package = 'BarnebyLives'),
  'rmarkdown', 'templates', 'labels', 'skeleton'
  )

# here we copy over one of the skeletons which we are going to modify in this example
file.copy(
  from = file.path(p2lib, 'skeleton-customfont.Rmd'), 
  to = file.path(local, 'skeleton-customfont.Rmd')
  )

rm(p2libs, folds)
```

If you look at the header on this file, you'll notice it has more going on than the other skeletons. 
As you will likely end up wanting to merge multiple label components together into a single skeleton (eventually), be sure to use this as the preamble. 

```
output: 
  pdf_document:
    latex_engine: xelatex # or lualatex
    keep_text: true
params:
  Collection_number: Collection_number
geometry: paperheight=4in,paperwidth=4in,margin=0.1in
header-includes:
  - \pagenumbering{gobble}
  - \usepackage{setspace}
  - \usepackage{fontspec}
  - \setmainfont{Georgia}
```
