---
title: "Custom Label Templates"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom Label Templates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Overview 
Obviously, everyone wants to be able to customize their label templates! 
Why go through the bother of installing and running `BarnebyLives` if you cannot do that! 
So here are a couple examples of tweaking around with the default skeleton to get some labels which will... Look familiar. 

Note that the labels are made using LaTeX, an awesome guide for LaTeX (relevant to the level you will be using it at) is [Overleaf.org](https://www.overleaf.com/learn/latex/Learn_LaTeX_in_30_minutes), and check the links in the side bar too! 

In this vignette, I will show you what the three default templates look like, and then we will make some slight customizations to them. 

## Examples 

As a reminder, the standard template looks like this. 

INSERT IMG

Another included template, for the Seeds of Success program, looks like this. 

INSERT IMG 

And our final default template looks like this

INSERT IMG

## Modifications

All of the labels can be copied from their original locations using the following code
```{r}
```

If you want to customize one, be sure to keep it in a 'local' location on your computer which will not be overwritten when you update the package, which may happen without your knowledge if this becomes a dependency to another package (which seems unlikely to me). 
Once the file is in that location it can easily be opened for safe editing. 

One final note is that LaTeX uses *spaces*, or ' ' for controlling content. 
In particular two spaces will force the text onto a new line. 
It is hard for me illustrate these in the examples, but I'll try and do my best to mention them when relevant. 

For this example we will focus on the final default template, which is a pretty middle of the road design. 

First let's change the line that controls the size of the font FOR ... 
```{r old font size}

```

Now we can change some alignment
```{r old alignment}

```

We may also decide that we don't care about presenting all of this information on the label
```{r old descriptions}

```

