---
title: "Custom label logos and maps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom label logos and maps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

## Overview 
Obviously, everyone wants to be able to customize their label templates! 
Why go through the bother of installing and running `BarnebyLives` if you cannot do that!  

This section is on embedding maps or logos onto labels. 

Note that the labels are made using LaTeX, an awesome guide for LaTeX (relevant to the level you will be using it at) is [Overleaf.org](https://www.overleaf.com/learn/latex/Learn_LaTeX_in_30_minutes), and check the links in the side bar too! 

This vignette assumes you were able to get through `Custom Label Templates`. 

## Prepare workspace

```{r load libraries and fine labels, eval =F}
library(BarnebyLives)
library(tidyverse)

local <- file.path('~', 'Documents', 'assoRted', 'Barneby_Lives_dev', 'LabelStyles')

l.nevadense <- collection_examples |>
  filter(Collection_number == 2823) |>
  mutate(
    Project_name = 'Custom Fonts', 
    Coordinate_uncertainty = '5m'
    ) 

write.csv(l.nevadense, file.path(local, 'SoS-ExampleCollection.csv'), row.names = F)
```

Label templates can be copied from their original locations using the following code.  

```{r copy a template to a local directory, eval = F}
p2lib <- file.path(
  system.file(package = 'BarnebyLives'),
  'rmarkdown', 'templates', 'labels', 'skeleton'
  )

# here we copy over one of the skeletons which we are going to modify in this example
file.copy(
  from = file.path(p2lib, 'skeleton-maps.Rmd'), 
  to = file.path(local, 'skeleton-maps.Rmd')
  )

# copy an image to use for a logo
file.copy(
  from = 
    file.path(
    system.file(package = 'BarnebyLives'), 
    'rmarkdown', 'SoS_transmittal', 'skeleton', 'logo.png'),
  to = file.path(local, 'logo.png')
  )

rm(p2lib)
```


We will also make a map to embed on a label. 

```{r, eval = F}
st <- tigris::states()

l.nevadense.sf <- sf::st_as_sf(
  l.nevadense, 
  coords = c('longitude_dd', 'latitude_dd'),
  crs = 4326
  )

# create a simple dot map of the state. 
map_maker(
  l.nevadense.sf, 
  st, 
  path = local,
  collection_col = 'Collection_number'
  )
```


## Render Labels 

```{r actually render labels, echo = F, eval = F}
purrr::walk(
  .x = l.nevadense$Collection_number,
  ~ rmarkdown::render(
    input = file.path(local, 'skeleton-maps.Rmd'), 
    output_file = file.path(local, paste0('custom-font', glue::glue("{.x}.pdf"))), 
    params = list(Collection_number = {.x}, imagepath = 'logo.png'), 
  )
)

purrr::walk(
  .x = l.nevadense$Collection_number,
  ~ rmarkdown::render(
    input = file.path(local, 'skeleton-customfont-twofonts.Rmd'), 
    output_file = file.path(local, paste0('custom-twofonts', glue::glue("{.x}.pdf"))), 
    params = list(
      Collection_number = {.x}, 
      imagepath = file.path('maps', 'map_2823.png')
      ) 
  )
)
```

