---
title: "Rendering a shipping manifest"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rendering a shipping manifest}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

I find it to be good practice to include both a shipping manifest and short note to the curators which are receiving material when I ship out my collection. 
The 'short notes' I generally send are slightly post-card-esque, unless working on some big government type project - which the note in the example will be. 
In general I just like these as a nice little way of communicating with someone with a little bit more personal touch. 

The shipping manifests are very important they allow me to check off material and make sure it is packed. 
They also seem like a cool old school way of helping people figure what is meant to be where.
More feel than *another* email.  
I've messed one up big time before, and well, they don't talk to me there anymore. 
So it's good to automate these I suppose. 

BarnebyLives has a default template for this process, which automatically combines the transmittal notice with the shipping manifest.  
As with the other templates from BL, it is easy to modify this template, and supply your local copy - with changes, in lieu of the default data set. 
Unlike the herbarium labels, even if you wanted to use this default template, you should copy this fill to a more user friendly location on your computer. 
You will need to manually update some values, and the file itself will be rendered to a pdf in the location which it is found. 
Here we copy the file over to a more 'natural' part of the computers file system just like we did with the label templates. 

```{r Create a local copy of transmittal notice, warning = F}
p2libs <- system.file(package = 'BarnebyLives')

folds <- file.path(
  'BarnebyLives', 'rmarkdown', 'templates', 'SoS_transmittal', 'skeleton', 'skeleton.Rmd')

file.copy(from = file.path(p2libs, folds), to =  '.')
```

Remember, changing the 'to' parameter in `file.copy` will allow it go wherever you want it go, in this example it is being copied into our local directory. 
Do not make changes to the file in the R directory (folder) holding the package itself - these changes will be overwritten and lost whenever you end up updating the package. 
Now you should have a skeleton for the transmittal notice. 
Open this up in a R tab, and fill out the `User` defined settings. 
An example is below (do this in the template! not here). 

```{r user defined settings}
herbarium_code <- 'COLO' # index herbariorum code
project <- data.frame(
  Name = 'Seeds of Success', # the project name. 
  Project_description = 'native seed conservation and development project', # short description
  Organization = 'Chicago Botanic Garden',
  Phone = '312-607-0257',
  Person = 'Reed Benkendorf',
  Email = 'rbenkendorf@chicagotbotanic.org',
  PersonTitle = 'Lead Botanist',
  Sets = 'quadruplicate',
  Permits = 'These materials were collected under the auspices of the US Department of Interiors Bureau of Land Management, unless noted otherwise.'
)
```

Once you have made all of your changes simply run the job. 
And once you've done that you can 'knit' the document using the 'knit' button in Rmarkdown's graphic user interface or `Ctrl + Shift + k`. 
