---
title: "running_pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{running_pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(tidyverse)
library(BarnebyLives)
```


## load data into R

### load data from a local (on computer) source

### load data through google sheets

An alternative format for loading data is through Google Sheets. 
This can facilitate data-sharing and standardization by many collectors in different locations. 
This is the source of data which was used during beta-testing for development. 

We will use googlesheets4, which is a part of the tidyverse, but I do not think installs by default with the remainder of the 'verse. 
We will feed in the last part of the URL as the link to the path. 
If this is your first time using googlesheets4, it will walk you through a few steps to log into your google account so you can access data securely. 


```{r load through google sheets, eF}

library(googlesheets4)
googledrive::drive_auth("reedbenkendorf27@gmail.com")
# read in data from the sheet to process
input <- read_sheet('1iOQBNeGqRJ3yhA-Sujas3xZ2Aw5rFkktUKv3N_e4o8M', 
                    sheet = 'Data Entry - Examples') %>% 
  mutate(UNIQUEID = paste0(Primary_Collector, Collection_number))

```


### only process data which have not yet run through the pipeline. 

```{r, eval = F}

# determine whether these data have already been processed by the script, using
# a unique combination of collection name and collection code. 
read_sheet('1iOQBNeGqRJ3yhA-Sujas3xZ2Aw5rFkktUKv3N_e4o8M',
                        sheet = 'Processed - Examples') %>% 
  select(Collection_number, Primary_Collector) %>% 
  mutate(UNIQUEID = paste0(Primary_Collector, Collection_number))

df <- filter(input, ! UNIQUEID %in% processed$UNIQUEID ) %>% 
  select(-UNIQUEID)

rm(processed, input)
```


## Ensure Coordinates and Dates and formatted appropriately

Date parsing, e.g. convert date into congruent museum formats (month in roman numerals, or spelled out).
This step assumes that you are recording your dates in the American format of (month/day/year), rather than the internationally more common format (day/month/year).

```{r Date parsing}
data <- date_parser(input, 'Date_digital')

dplyr::select(data, starts_with('Date')) %>% 
  utils::head()
```

Conversion of Degrees Minutes Seconds (DMS) to Decimal Degrees (DD). 
The DMS format is still used in a few applications, such as the Seeds of Success program, we will convert this to the more generally utilized DD format. 

```{r Conversion of Degrees Minutes Seconds (DMS) to Decimal Degrees (DD)}
data <- dms2dd(data, dms = F)

dplyr::select(data, starts_with(c('latitude', 'longitude'))) %>% 
  utils::head()
```

For internal usage and data acquisition we will make data set explicitly spatial, rather than just holding coordinates in a column. 
This is also required to write out data in a format readable by Google Earth or another Geographic Information System. 

```{r}

data <- coords2sf(data)

head(data) # now we can see that it is an sf object
```


## Retrieve Political/Administrative Information

BarnebyLives! will retrieve various components of both political and administrative information. All of these data which include: The *state*, and *county* of collection, and the *township, range, section (TRS)* from the Public Land Survey System (PLSS). 
It will also determine whether a Federal or State Agency manages or owns the land which were collected from, and if so return the name of it. 
If either the United States Forest Service or Bureau of Land Management does manage the land, it will return information on grazing allotments.

Retrieve State, County, Ownership, TRS, and allotment information
```{r Retrieve Political, eval = F}
political_grabber(data)
```

## Geographic

Retrieve Nearest GNIS place name, and azimuth from it
```{r Retrieve Place and location from it, eval = F}
data <- site_writer(data)
```

## Site characteristics

Elevation (both meters and feet), Slope, Aspect, Surficial geology, and geomorphons

```{r, eval = F}

physical_grabber()

```


## Taxonomic

BarnebyLives performs multiple operations related to taxonomy and nomenclature. The function **spell_check()**, will attempt to determine whether the supplied binomial scientific name, and family are spelled correctly and offer suggestions if they are not. 
This function requires that the first few letters of the generic name and epithet are spelled correctly. It will run a spell check against nearly all published plant name spellings, and search for the most similar spelling within these lists if material is not found. 
It will operate on both pieces of the binomial separately, and can work to identify the species within the context of genus. 

```{r check spelling of binomial and family, eval = F}
spell_check()
```

A similar check is to determine that the author whom is cited for describing the species, or transferring it to another genus, is spelled using the standard author abbreviations.
This check operates under a similar manner to **spell_check()**.
The list which is consults *is* comprehensive, within a few years, of all taxonomists whom have published plant species. 

```{r check spelling of author, eval = F}

```

After BarnebyLives! has verified the spelling of the species it will submit it to Plants of the World Online to determine whether a more current and accepted name has been applied. 
BL! will not automatically over-write your submitted species, and it is important that you interpret the results of query.
Especially as it is possible that **spell_check** has mis-matched your material to another name. 

Searches for synonym to species
```{r Search for synonyms to species, eval = F}
powo_seacher()
```

The final process in the taxonomic step is to ensure that the spellings of any species noted in the *vegetation*, or *associates* fields are accurate. These *will not* then be submitted to Kew, as the fields could quickly overwhelm the API. 

```{r check spellings of vegetation and associated species, eval = F}

```

## Directions

directions to a parking spot can be acquired from Google Maps; however this implies the location can be driven to in the first place. 
This will require interactive alterations from the user.

```{r}
library(googleway)
SoS_gkey = Sys.getenv("Sos_gkey")

test_gq <- get_gooogle_directions(data, api_key = SoS_gkey)

# the locations at which a turn occurs
#turns <- googlePolylines::decode( df$routes$overview_polyline$points ) %>% 
#  bind_rows() %>% 
#  sf::st_as_sf(coords = c('lon', 'lat'), crs = 4326)

```


## Export collections as 'shapefile' or KML for use in GIS or GoogleEarth

```{r write out data to googlesheets4}

# we will add these data onto our final sheet.  THIS IS THE LAST STEP OF THIS TUTORIAL
sheet_append('1iOQBNeGqRJ3yhA-Sujas3xZ2Aw5rFkktUKv3N_e4o8M', 
             sheet = 'Processed - Examples', data = df) 

```


