---
title: "running_pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{running_pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(BarnebyLives)
```


## Ensure Coordinates and Dates and formatted appropriately

Date parsing, e.g. convert date into congruent museum formats (month in roman numerals)
```{r Date parsing}

date_parser()
```

Conversion of Degrees Minutes Seconds (DMS) to Decimal Degrees (DD)
```{r Conversion of Degrees Minutes Seconds (DMS) to Decimal Degrees (DD)}

dms2dd()
```


Make dataset explicitly spatial
```{r}
coords2sf()
```



## Retrieve Political Information

Retrieve State, County, Ownership, TRS, and allotment information
```{r Retrieve Political}

political_grabber()

```




## Geographic

Retrieve Mountain Range (if relevant)
```{r Retrieve Mountain Range}

```

Retrieve Nearest GNIS place name, and azimuth from it
```{r Retrieve Place and location from it }

```

## Site characteristics

Elevation (both meters and feet), Slope, Aspect, Surficial geology, and geomorphons

```{r}

physical_grabber()

```


## Taxonomic


Quick spell check for Family, Genus, and specific epithet
```{r}
spell_check()
```

Consults taxonomic authors abbreviations to ensure correct spelling
```{r}

```

Searches for synonym to species
```{r}
powo_seacher()
```


## Directions

directions to a parking spot can be acquired from Google Maps; however this implies the location can be driven to in the first place. This will require interactive alterations from the user.

## Export collections as 'shapefile' or KML for use in GIS or GoogleEarth

```{r}

```


