---
title: "running pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{running pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Purpose 
The purpose of this script is to assist botanical collectors in accessioning their collections. 
It is the central component of this entire package. 
Here we will go over ensuring that values which will be printed onto herbarium labels and uploaded into databases are accurate. 

It is our experience that many herbaria now request digital copies of collection information in their own format. 
This script aims to alleviate the investment in time this requires, and in doing so allow the collector more time to do less onerous tasks; such as collection and study. 

While in theory all of the tasks in this script could be automated, doing so would be an unconscionable action against local taxonomic authorities. 
While several database options for updating collections (and spell checking names) which are current and well curated (e.g. Kew Plants of the World) exist, we feel many proposals and combinations require interpretation by those in the regions attempting to understand the material at hand. 

This script will perform the following tasks.  

### Geographic

**1)** Convert coordinates to a Decimal Degrees format, if needed (e.g. from DMS 72*17.55 -> 72.29861).  
**2)** Retrieve appropriate political place names. (i.e. State, County).  
**3)** Retrieve accurate elevation values of coordinates.  
**4)** Retrieve Public Land Ownership (e.g. United States Forest Service, Bureau of Land Management).  
**5)** Project coordinates from one system to another, if needed (e.g. from an UTM zone to WGS84).  
**6)** Retrieve Cadastral data from the PLSS for Township, Range, Section (TRS)  

### Temporal

**1)** Provide a 'written' date in the style of choosing. (e.g. 12th May 2021), for the purpose of labeling.

### Taxonomic

**1)** Provide a 'spell check' for family.  
**2)** Provide a 'spell check' for genus.  
**3)** Provide a 'spell check' for epithet.  
**4)** Provide a 'spell check' for authorities.  
**5)** In addition you may check for synonymy at each of these taxonomic levels, and accept or reject suggestions.  

### Directions  

I preface this with the observation that a good deal of collectors and curators consider these data non-essential now. Regardless, if you are so inclined to gain access to a free google maps account these services are provided. 

**1)** Write directions, from a town of notable size or repute, to the closet parking spot to the collection locality.

Note that these directions may not reflect the actual route traveled by the collectors, but rather what Google Maps thinks is optimal - at the time of the query. *Caveat emptor.*

# Usage

Files required to use this script:

- PAD-US database for retrieving land ownership information if needed (*note does not include identities of Private owners*). 
- World Flora Online static copy of the database.

```{r setup, message = FALSE}
# devtools::install_github('sagesteppe/BarnebyLives')
library(tidyverse)
library(BarnebyLives)
```

In this example I am going to showcase loading the data we want to run through the pipeline using Google Sheets. 
This can allow for collectors on a team to readily contribute to preparing and transcribing notes for an expedition simultaneously. 
Loading data in from a local CSV or excel type file is as easy as `read.csv` or `readxl::read_excel` instead. 

We will use the package `googlesheets4`, which is a part of the tidyverse, but which does not install by default with the remainder of the 'verse. 
We will feed in the last part of the URL as the link to the path. 
If this is your first time using `googlesheets4`, it will walk you through a few steps to log into your google account so you can access data securely. 
After that `googlesheets4` is actually really straightforward to use, note that I use the `drive_auth` function here. 
This function makes non-interactive use easy, it is only used here because the vignette renders as a non-interactive format. 

```{r load through google sheets, eval = F}
# install.packages('googlesheets4')
library(googlesheets4)

googledrive::drive_auth("reedbenkendorf27@gmail.com")
# read in data from the sheet to process
df <- read_sheet('1iOQBNeGqRJ3yhA-Sujas3xZ2Aw5rFkktUKv3N_e4o8M', 
                    sheet = 'Data Entry - Examples') %>% 
  mutate(UNIQUEID = paste0(Primary_Collector, Collection_number))

```

### only process data which have not yet run through the pipeline. 

BarnebyLives takes a little bit of time to run, and it also requires the usage of Google Developer credits for the directions portion. 
You may want to ensure that you are not re-running files which you have already processed. 
We can do that very simply using a combination of collectors name's and their collection numbers. 
If you are processing for many many people at once, and worried people may have the same names, and be collecting in the same range of numbers, then you can add in other data which would be required to make a UNIQUEID, most likely a component of the date. 

```{r determine which data has been processed, eval = F}

# determine whether these data have already been processed by the script, using
# a unique combination of collection name and collection code. 
# these are the data which have gone through the pipeline already
processed <- read_sheet('1iOQBNeGqRJ3yhA-Sujas3xZ2Aw5rFkktUKv3N_e4o8M',
                        sheet = 'Processed - Examples') %>% 
  select(Collection_number, Primary_Collector) %>% 
  mutate(UNIQUEID = paste0(Primary_Collector, Collection_number))

# here we look for any 'input' files which are present in the processed files and 
# drop them. 
input <- filter(df, ! UNIQUEID %in% processed$UNIQUEID ) %>% 
  select(-UNIQUEID)

rm(processed, df)
```

Even though I coded this functionality, I essentially never use it. 
The reason being that I don't use the directions that much, and just get up to start a cup of tea when the pipeline is running. 

## Ensure Coordinates and Dates and formatted appropriately

Date parsing, e.g. convert date into congruent museum formats (month in roman numerals, or spelled out).
This step assumes that you are recording your dates in the American format of (month/day/year), rather than the internationally more common format (day/month/year).
We will also parse out the day of month, month of year, and year, which is used in multiple possible downstream software.

```{r Date parsing, eval = F}
data <- date_parser(input, coll_date = 'Date_digital', det_date = 'Determined_date')

dplyr::select(data, starts_with('Date')) %>% 
  utils::head()
```

### Conversion of Degrees Minutes Seconds (DMS) to Decimal Degrees (DD). 
The DMS format is still used in a few applications we will convert this to the more generally utilized DD format. 
Although DMS is certainly prettier to read, DD is just used so much more often we change everything to it. Sad. 
We will maintain both formats in the data set, so others have them readily available for other uses aside from creating herbarium labels. 

```{r Conversion of Degrees Minutes Seconds (DMS) to Decimal Degrees (DD), eval = F}
data <- dms2dd(data, dms = F)

dplyr::select(data, starts_with(c('latitude', 'longitude'))) %>% 
  utils::head()
```

Several spreadsheet software auto-increment numeric columns by default.
We try to identify that issue here. 
```{r check for autofill on coords, eval = F}
data <- autofill_checker(data)

dplyr::select(data, Long_AutoFill_Flag, Lat_AutoFill_Flag) |> 
  dplyr::drop_na() |>
  utils::head()
```

For internal usage and data acquisition we will make data set explicitly spatial, rather than just holding coordinates in a column. 
This is also required to write out data in a format readable by Google Earth or another Geographic Information System. 

```{r Make dataset explicitly spatial, eval = F}
data <- coords2sf(data)
head(data) # now we can see that it is an sf object
```

## Retrieve Political/Administrative Information

BarnebyLives! will retrieve various components of both political and administrative information. All of these data which include: The *state*, and *county* of collection, and the *township, range, section (TRS)* from the Public Land Survey System (PLSS). 
It will also determine whether a Federal or State Agency manages or owns the land which were collected from, and if so return the name of it. 
If either the United States Forest Service or Bureau of Land Management does manage the land, it will return information on grazing allotments.

Retrieve State, County, Ownership, TRS, and allotment information
```{r Retrieve Political, eval = F}
p <- '/media/steppe/hdd/Barneby_Lives-dev/geodata'

data <- political_grabber(data, y = 'Collection_number', path = p)
```

## Geographic Information for Locality info

Many younger collectors use Site Names which are quite broad, we advocate for the usage of locality names. 
A locality name is going to provide the relation between a site and the actual collection area through a distance and azimuth.
All of the sites which we use as prospective identifiers are from the Geographic Names Information System, these names have undergone standardization, and general cleaning up to better reflect currently acceptable naming practices (i.e. misogynistic and racist names have largely - if not entirely - been removed). 
Using GNIS allows for places to be unambiguously related to each other. 
While I hate saying that a collection site is XX km from a river at 121 degrees, because well - just think about it, that seems to be what people want. 

Retrieve Nearest GNIS place name, and azimuth from it
```{r Retrieve Place and location from it, eval = F}
data <- site_writer(data, path = p)
```

## Site characteristics

Elevation (both meters and feet), Slope, Aspect, Surficial geology, and geomorphons

```{r, eval = F}
data <- physical_grabber(data, path = p)
```

## Taxonomic

BarnebyLives performs multiple operations related to taxonomy and nomenclature.
The function **spell_check()**, will attempt to determine whether the supplied binomial scientific name, and family are spelled correctly and offer suggestions if they are not. 

This function requires that the first few letters of the generic name and epithet are spelled correctly. 
It will run a spell check against nearly all published plant name spellings, and search for the most similar spelling within these lists if material is not found. 
It will operate on both pieces of the binomial separately, and can work to identify the species within the context of genus. 

```{r check spelling of binomial and family, eval = F}
p <- '/media/steppe/hdd/Barneby_Lives-dev/taxonomic_data'
data <- spell_check(data, path = p)
```

A similar check is to determine that the author whom is cited for describing the species, or transferring it to another genus, is spelled using the standard author abbreviations.
This check operates under a similar manner to **spell_check()**.
The list which is consults *is* comprehensive, within a few years, of all taxonomists whom have published plant species. 

```{r check spelling of author, eval = F}
data <- author_check(data, path = p)
```

After BarnebyLives! has verified the spelling of the species it will submit it to Plants of the World Online to determine whether a more current and accepted name has been applied. 
BL! will not automatically over-write your submitted species, and it is important that you interpret the results of query.
Especially as it is possible that **spell_check** has mis-matched your material to another name. 

Searches for synonym to species
```{r Search for synonyms to species, eval = F}

names <- sf::st_drop_geometry(data) %>% 
  pull(Full_name)

pow_res <- lapply(names,
      powo_searcher) %>% 
  bind_rows()
data <- bind_cols(data, pow_res)

rm(names, pow_res)
```

The final process in the taxonomic step is to ensure that the spellings of any species noted in the *vegetation*, or *associates* fields are accurate. These *will not* then be submitted to Kew, as the fields could quickly overwhelm the API. 

```{r check spellings of vegetation and associated species, eval = F}

```

Remove the collected species from the list of associated species. Many collectors will include the focal collection in 
the list of associated species when performing data entry for a site. We can remove it here.
```{r drop focal species from associated species, eval = F}
data <- associate_dropper(data)
```


## Directions

directions to a parking spot can be acquired from Google Maps; however this implies the location can be driven to in the first place. 
This will require interactive alterations from the user.

```{r site directions, eval = F}
SoS_gkey = Sys.getenv("Sos_gkey")
directions <- directions_grabber(data, api_key = SoS_gkey)
```


## Export collections

We will write these data to Google sheets for our examples. 

```{r write out data to googlesheets4, eval = F}

# first ensure the columns are in the same order as google sheets

#processed <- read_sheet('1iOQBNeGqRJ3yhA-Sujas3xZ2Aw5rFkktUKv3N_e4o8M',
#                        sheet = 'Processed - Examples') %>% 
#  colnames() 

df <- sf::st_drop_geometry(directions) #%>% 
#  select(all_of(processed))
# we will add these data onto our final sheet.
sheet_append('1iOQBNeGqRJ3yhA-Sujas3xZ2Aw5rFkktUKv3N_e4o8M', 
             sheet = 'Processed - Examples', data = df) 

```

We can also export collections as either (or both!) a 'shapefile' or KML for use in GIS or GoogleEarth.

```{r write out collections in a spatial data format, eval = F}

geodata_writer(data, path = 'someplace', 
               filename = 'Herbarium_Collections_2023',
               filetype = 'kml')

```


