---
title: "setting up files"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{setting up files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, eval = F, message = F}
remotes::install_github('sagesteppe/BarnebyLives')
library(BarnebyLives)
library(tidyverse)
```

BarnebyLives relies on a variety of data to be downloaded locally to your computer to work. 
My instance, which covers a large area, essentially from Michigan to California, South through Texas to the Canadian border is around 16GiB. 
Many more regional collections are likely to be in the realm of 4-10 GiB, so relative to a normal R package A LOT of data. 
Relative to a typical geospatial workflow.... Essentially no data.
Given how much larger the amount of disk data are on computers I think this amount is trivial to consider. 

It's important that you decide where BarnebyLives will be installed on your computer. 
While I have couple computers, I maintain only a single instance of BarnebyLives. 
It lives on one of my 3.5 terabyte hard drives which I have set up for various science projects. 

I will be downloading and copying some of the data to that location, where I have a directory named 'Barneby_Lives-dev'. 

## Taxonomic Data

**Worldwide Checklist of Vascular Plants** provides the local taxonomic information required to perform much of the name matching.
We utilize it to ensure that queries sent to **Plants of The World Online** are spelled correctly, in order to reduce the number of queries we make to it. 
"The World Checklist of Vascular Plants, a continuously updated resource for exploring global plant diversity" serves as the basis for Kews Plants of the Worlds, this publication details the process behind the data set. 
We can download a static copy of the data from Kews data portal at: http://sftp.kew.org/pub/data-repositories/WCVP/, I simply download the data manually rather than by script. 
This is not that large of a file, so it be downloaded in a couple minutes. 

We subset names by continent, I have experimented using Regsions within continents, but the results are usually wanting. 
```{r read in WCVP data, eval = F}
p2 <- '/media/steppe/hdd/Barneby_Lives-dev/taxonomic_data' # note we have downloaded it 
# to this location - this is the location where our taxonomy data will be unpacked to!

TaxUnpack(path = p2, continents = 'NORTHERN AMERICA')
```

The way that BarnebyLives matches spellings of names is very simple. 
It creates three tables, one large table which contains all of the species in the WFO database, one moderately small table which contains all infraspecific taxa (varieties and subspecies), and a small table of all of the genera. 

However, both the genus and specific epithet tables are abbreviated. 
We find the results from these tables based on only the leading first two or three characters. 
This gives rises to one of BarnebyLives(!) limitations, the assumption that the first two characters of the genus, and first three letters of the specific epithet are spelled correctly. 

Now we will also prepare the author abbreviations data set. 
These data come from [IPNI](https://www.ipni.org/)

```{r Prepare material for authorship lookup tables, eval = F}

abbrevs <- read.csv(
  file.path(
    system.file(package = 'BarnebyLives'), 
    'data-raw', 'ipni_author_abbreviations.csv')
)

trailed <- vector(mode = 'character', length = nrow(abbrevs))
trailed[grep('\\.$', abbrevs$author_abbrevation)] <- '.'
 # identify which abbreviations have a trailing period

abbrevs_spaced <- sub('\\.$', '', abbrevs$author_abbrevation) # remove the trailing periods
abbrevs_notrail <- sub('\\.(?!.*\\.)', ". ", 
    abbrevs_spaced, perl = T) # identify the last period in the name, and add a space after it

abbrevs <- paste0(abbrevs_notrail, t)
write.csv(abbrevs, row.names = F, file.path(p2, 'abbrevs.csv'))

rm(trailed, abbrevs_notrail, abbrevs_spaced, abbrevs)
```


## Geographic data

Many of these data sets are very large we are going to crop them to the area of analysis, and save the subsets. This package depends on **terra** and makes use of Virtual Raster Tiles to query parameters. 

```{r library, message = F}
library(sf)
library(terra)
```

You can readily define a simple custom extent for your study area. The coordinates are fine to grab and round from Google Maps or similar.
What is important is that the 5th set of coordinates is a repeat of the 1st set of coordinates. 

```{r define domain of data}

bound <- data.frame(
  y = c(30, 30, 50, 50, 30), 
  x = c(-85, -125, -125, -85, -85)
) %>% 
  st_as_sf(coords = c('x', 'y'), crs = 4326) %>% 
  st_bbox() %>% 
  st_as_sfc()

bb_vals <- c(-125, -85, 30, 50)
w_illi <- ext(bb_vals)
```

The data which we obtain our data sources from come as very large tiles. We will clip these down to the extent of our analysis. Essentially raster tiles should be no larger than 2 gigabytes. So we need to ensure that we create a tile grid which will allow us to save these data under those constraints. 

```{r create tile template grid}
tile_cells <- st_make_grid(
  bound,
  what = "polygons", square = T, flat_topped = F, crs = 4326,
  n = c(
  abs(round( (bb_vals[2] - bb_vals[1]) / 5, 0)), # rows
    abs(round( (bb_vals[3] - bb_vals[4]) / 5, 0))) #cols
) %>% 
  st_as_sf() %>% 
  rename(geometry = x) %>% 
  mutate(x = st_coordinates(st_centroid(.))[,1],
         y = st_coordinates(st_centroid(.))[,2], 
         .before = geometry, 
         across(c('x', 'y'), \(x) round(x, 1)),
         cellname = paste0('n', abs(y), 'w', abs(x))
         )

tile_cellsV <- vect(tile_cells)
```

This package comes with the 'mason' function, which will cut up the raster data sets, and save them in a new location. 

The source for our physical raster data is 'Geomorpho90m, empirical evaluation and accuracy assessment of global high-resolution geomorphometric layers'.
These data are available from here https://opentopography.s3.sdsc.edu/minio/dataspace/OTDS.012020.4326.1/raster/

```{r subset physical variables to domain, eval = F}

mason(dirin = '../geodata_raw/geomorphons/',
      dirout = '../geodata/geomorphons/',
      grid = tile_cellsV, fnames = 'cellname')

mason(dirin = '../geodata_raw/aspect/',
      dirout = '../geodata/aspect/',
      grid = tile_cellsV, fnames = 'cellname')

mason(dirin = '../geodata_raw/slope/',
      dirout = '../geodata/slope/',
      grid = tile_cellsV, fnames = 'cellname')
      
mason(dirin = '../geodata_raw/elevation/',
      dirout = '../geodata/elevation/',
      grid = tile_cellsV, fnames = 'cellname')

rm(mason)
```

We will simply crop the vector data to the limit of the study area. We utilize the U.S. Census Bureau's data for this. We can download the States data set from the U.S. Census Bureau's website. The other data sets can be loaded in from the tigris r package. 

```{r crop political boundary data to extent, eval = F}
counties <- st_read('../geodata_raw/USCounties/tl_2020_us_county.shp', quiet = T) %>% 
  st_transform(st_crs(tile_cells)) %>% 
  select(STATEFP, COUNTY = NAME)

states <- tigris::states() %>% 
  select(STATEFP, STATE = NAME, STUSPS) %>% 
  st_drop_geometry()

counties <- counties[st_intersects(counties, st_union(tile_cells)) %>% lengths > 0, ]
counties <- left_join(counties, states, by = "STATEFP") %>% 
  select(-STATEFP) %>% 
  arrange(STATE)

states <- counties %>%
  distinct(STUSPS) %>% 
  pull(STUSPS)

st_write(counties, dsn = file.path('../geodata/political', 'political.shp'))

rm(counties)
```

We will set up our places data set which we will use go gather driving directions from Google maps. 
```{r, eval = F}

bb <- pgirmess::bbox2sf(w = bb_vals[1], e = bb_vals[2],
                        s = bb_vals[3], n = bb_vals[4]) %>% 
  st_as_sf()

st <- tigris::states() %>% 
  st_transform(4326)
st <- st[st_intersects(st, bb) %>% lengths > 0, c('STATEFP', 'NAME', 'STUSPS') ]
places <- lapply(st$STUSPS, tigris::places)

places <- places %>% 
  bind_rows() %>% 
  filter(str_detect(NAMELSAD, 'city')) %>% 
  dplyr::select(STATEFP, NAME) %>% 
  sf::st_centroid() %>% 
  st_transform(4326) 

places <- places %>% 
  rename(CITY = NAME) %>% 
  left_join(., st %>% 
              st_drop_geometry(), by = 'STATEFP') %>% 
  st_transform(4326)

places <- places[st_intersects(places, bb) %>% lengths > 0, ]
places <- st_set_precision(places, precision=10^3)
st_write(places, 'places.shp', append = F) # this will drop location coordinates to 3 decimal places. 
places <- st_read('places.shp')

google_towns <- select(places, -STUSPS, -STATEFP, STATE = NAME) %>% 
  rowid_to_column("ID")

usethis::use_data(google_towns)

rm(places, google_towns)
```


We can access a Mountains dataset from EarthEnv which we can use for portions of our place names here https://www.earthenv.org/mountains.

```{r crop mountain names database to extent, eval = F}

mtns <- st_read('../geodata_raw/globalMountains/GMBA_Inventory_v2.0_standard_basic.shp', 
        quiet = T) %>% 
  st_make_valid() %>% 
  select(MapName) %>% 
  st_crop(., st_union(tile_cells))|>
  dplyr::rename(Mountains = MapName) |>
  dplyr::mutate(Mountains = stringr::str_remove(Mountains, '[(]nn[])]'))

st_write(mtns, dsn = file.path('../geodata/mountains', 'mountains.shp'), quiet = T)

rm(mtns)
```

Our place names are acquired from the United States Geological Surveys GNIS data set https://www.usgs.gov/tools/geographic-names-information-system-gnis .

```{r crop geographic place name database to extent, eval = F}

files <- paste0('../geodata_raw/GNIS/Text/DomesticNames_', states, '.txt')
cols <- c('feature_name', 'prim_lat_dec', 'prim_long_dec')

places <- lapply(files, read.csv, sep = "|") %>% 
  map(., ~ .x |> select(all_of(cols))) %>% 
  data.table::rbindlist() %>% 
  st_as_sf(coords = c('prim_long_dec', 'prim_lat_dec'), crs = 4269) %>% 
  st_transform(4326)

places <- places[st_intersects(places, st_union(tile_cells)) %>% lengths > 0, ]
places <- places %>% 
  mutate(ID = 1:nrow(.))

st_write(places, dsn = file.path('../geodata/places', 'places.shp'), quiet = T)

rm(places, cols, files)
```

To determine the land ownership of collections we utilize the Protected Areas Database-United States also from the USGS  https://www.usgs.gov/programs/gap-analysis-project/science/pad-us-data-download .

```{r crop padus to domain, eval = F}

padus <- st_read(dsn = '../geodata_raw/PADUS3/PAD_US3_0.gdb', 
                 layer = 'PADUS3_0Fee' %>% 
  filter(State_Nm %in% states) %>% 
  select(Mang_Name, Unit_Nm) %>% 
  st_cast('MULTIPOLYGON')

tile_cells <- st_transform(tile_cells, st_crs(padus))
padus <- padus[st_intersects(padus, st_union(tile_cells)) %>% lengths > 0, ]
padus <- st_transform(padus, crs = 4326)

st_write(padus, dsn = file.path('../geodata/pad', 'pad.shp'), quiet = T)

# replace really long state land board names with 'STATE SLB'

padus <- st_read('../../Barneby_Lives-dev/geodata/pad/pad.shp')

padus <- padus %>% 
  mutate(Unit_Nm = 
           str_replace(Unit_Nm, 
                               'The State of Utah School and Institutional Trust Lands Administration', 'Utah')
         )

st_write(padus, dsn = file.path('../geodata/pad', 'pad.shp'), quiet = T, append = F)

rm(padus)
```

The best geological data set are also, naturally, provided by the USGS. The geodatabase can be downloaded from here https://mrdata.usgs.gov/geology/state/ 

```{r geological map, eval = F}
geological <- st_read(
  '../geodata_raw/geologicalMap/USGS_StateGeologicMapCompilation_ver1.1.gdb', 
          layer = 'SGMC_Geology', quiet = T) %>% 
  select(GENERALIZED_LITH, UNIT_NAME) 

tile_cells <- st_transform(tile_cells, st_crs(geological))
geological <- geological[st_intersects(geological, st_union(tile_cells)) %>% lengths > 0, ]
geological <- st_crop(geological, st_union(tile_cells))
geological <- st_transform(geological, 4326)

st_write(geological, dsn = file.path('../geodata/geology', 'geology.shp'), quiet = T)

tile_cells <- st_transform(tile_cells, crs = 4326)
rm(geological)
```

Grazing allotments for the bureau of land management can be downloaded from https://gbp-blm-egis.hub.arcgis.com/datasets/blm-natl-grazing-allotment-polygons 

Grazing allotments for the United States Forest Service can be downloaded from ...
 
```{r allotments, eval = F}

blm <- st_read(
  '../geodata_raw/BLMAllotments/BLM_Natl_Grazing_Allotment_Polygons.shp') %>% 
  select(ALLOT_NAME)
usfs <- st_read(
  '../geodata_raw/USFSAllotment/S_USA.Allotment.shp', quiet = T
) %>% 
  select(ALLOT_NAME = ALLOTMENT_) 

allotments <- bind_rows(blm, usfs) %>% 
  st_make_valid()
tile_cells <- st_transform(tile_cells, st_crs(allotments))
allotments <- allotments[st_intersects(allotments, st_union(tile_cells)) %>% lengths > 0, ]
allotments <- st_crop(allotments, st_union(tile_cells))
allotments <- st_transform(allotments, 4326)

st_write(allotments, dsn = file.path('../geodata/allotments', 'allotments.shp'), quiet = T)

rm(blm, usfs, allotments)
```



```{r public land survey system only for land west of CO border, eval = F}

bound_nm <- data.frame(
  y = c(31, 31, 50, 50, 31), 
  x = c(-95, -125, -125, -95, -95)
) %>% 
  st_as_sf(coords = c('x', 'y'), crs = 4326) %>% 
  st_bbox() %>% 
  st_as_sfc()

township <- st_read(
  '../geodata_raw/nationalPLSS/ilmocplss.gdb', layer = 'PLSSTownship', quiet = T
) %>% 
  st_cast('POLYGON') %>% 
  select(TWNSHPLAB, PLSSID)

bound_nm <- st_transform(bound_nm, st_crs(township))
township <- township[st_intersects(township, bound_nm) %>% lengths > 0, ]
township <- st_drop_geometry(township)

section <- st_read(
  '../geodata_raw/nationalPLSS/ilmocplss.gdb', layer = 'PLSSFirstDivision', quiet = T
) %>% 
  st_cast('POLYGON') %>% 
  select(FRSTDIVLAB, PLSSID, FRSTDIVID)

section <- section[st_intersects(section, bound_nm) %>% lengths > 0, ]

trs <- inner_join(section, township, 'PLSSID') %>% 
  mutate(TRS = paste(TWNSHPLAB, FRSTDIVLAB)) %>% 
  select(TRS)

before <- object.size(trs)
trs <- st_simplify(trs)
after <- object.size(trs)
object.size(trs)

st_write(trs, 
         dsn = file.path('../geodata/plss', 'plss.shp'), quiet = T)

rm(section, township, trs, bound_nm, before, after)
```

```{r modify names of variables to be in proper formats, eval = F}

sf::st_read('../geodata/political/political.shp', quiet = T) %>% 
  rename(County = COUNTY, State = STATE) %>% 
  st_write(., '../geodata/political/political.shp', quiet = T, append = F)

sf::st_read('../geodata/allotments/allotments.shp', quiet = T) %>% 
  dplyr::mutate(Allotment = str_to_title(ALLOT_NAME)) %>% 
  dplyr::select(-ALLOT_NAME) %>% 
  st_write(., '../geodata/allotments/allotments.shp', quiet = T, append = F)

sf::st_read('../geodata/plss/plss.shp', quiet = T) %>% 
  rename(trs = TRS) %>% 
  st_write(., '../geodata/plss/plss.shp', quiet = T, append = F)

sf::st_read('../geodata/pad/pad.shp', quiet = T) %>% 
  sf::st_make_valid() %>% 
  dplyr::filter(sf::st_is_valid(.)) %>% 
  mutate(Unit_Nm = str_replace(Unit_Nm, "National Forest", "NF"),
         Unit_Nm = str_replace(Unit_Nm, "District Office", "DO"), 
         Unit_Nm = str_replace(Unit_Nm, "Field Office", "FO"), 
         Unit_Nm = str_replace(Unit_Nm, "National Park", "NP"),
         Unit_Nm = str_replace(Unit_Nm, 'National Grassland', 'NG'),
         Unit_Nm = str_replace(Unit_Nm, "National Wildlife Refuge|National Wildlife Range", "NWR"),
         Unit_Nm = str_replace(Unit_Nm, 'National Monument', 'NM'),
         Unit_Nm = str_replace(Unit_Nm, "National Recreation Area", "NRA"), 
         Unit_Nm = str_replace(Unit_Nm, "State Resource Management Area", "SRMA"), 
         Unit_Nm = str_replace(Unit_Nm, "State Trust Lands", "STL"), 
         Unit_Nm = str_replace(Unit_Nm, "Department", "Dept."),
         Unit_Nm = str_replace(Unit_Nm, "Recreation Area", "Rec. Area"),
         Unit_Nm = str_replace(Unit_Nm, 'Wildlife Habitat Management Area', "WHMA"), 
         Unit_Nm = str_replace(Unit_Nm, 'Wildlife Management Area', "WMA")) %>% 
  st_write(., '../geodata/pad/pad.shp', quiet = T, append = F)


sf::st_read('../geodata/places/places.shp', quiet = T) %>% 
  mutate(
    fetr_nm = str_remove(fetr_nm, ' Census Designated Place' ),
    fetr_nm = str_remove(fetr_nm, 'Township of '), 
    fetr_nm = str_remove(fetr_nm, ' \\(historical\\)'),
    fetr_nm = str_remove(fetr_nm, 'Village of '),
    fetr_nm = str_remove(fetr_nm, 'Town of '), 
    fetr_nm = str_remove(fetr_nm, 'City of ')
    ) %>% 
  filter(!str_detect(fetr_nm,
                     'Election Precinct| County|Ditch|Unorganized Territory|Canal|Lateral|Division|
                     River Division|Drain Number|Waste Pond|Commissioner District|Lake Superior|Lake Michigan')) %>% 
  mutate(
    fetr_nm = str_replace(fetr_nm, ' Canyon', ' cnyn.'),
    fetr_nm = str_replace(fetr_nm, ' Campground', ' cmpgrnd.'),
    fetr_nm = str_replace(fetr_nm, ' Creek', ' crk.'),
    fetr_nm = str_replace(fetr_nm, ' Fork', ' fk.'),
    fetr_nm = str_replace(fetr_nm, ' Lake', ' lk.'),
    fetr_nm = str_replace(fetr_nm, ' Meadows', ' mdws.'),
    fetr_nm = str_replace(fetr_nm, ' Meadow', ' mdw.'),
    fetr_nm = str_replace(fetr_nm, ' Mountains', ' mtns.'),
    fetr_nm = str_replace(fetr_nm, ' Mountain', ' mtn.'),
    fetr_nm = str_replace(fetr_nm, ' Number', ' #'),
    fetr_nm = str_replace(fetr_nm, ' Point', ' pt.'),
    fetr_nm = str_replace(fetr_nm, ' Reservoir', ' rsvr.'),
    fetr_nm = str_replace(fetr_nm, ' River ', ' rvr.')
  ) %>% 
  st_write(., '../geodata/places/places.shp', quiet = T, append = F)

```


# add files for templates

```{r, eval = F}
herbaria_info <- read.csv('/media/sagesteppe/ExternalHD/Barneby_Lives-dev/herbarium_data/index_herbariorum_info.csv')
usethis::use_data(herbaria_info)
```

```{r, eval = F}
collection_examples <- read.csv('/media/sagesteppe/ExternalHD/Barneby_Lives-dev/herbarium_data/shipping_examples.csv')
collection_examples[collection_examples==""] <- NA
usethis::use_data(collection_examples, overwrite = TRUE)
```


```{r, eval = F}
project_examples <- 
  read.csv('/media/sagesteppe/ExternalHD/Barneby_Lives-dev/herbarium_data/project.csv')
usethis::use_data(project_examples)
```

# Save the output formats for different herbaria databases

```{r, eval = F}
database_templates <- 
  read.csv('/media/steppe/hdd/Barneby_Lives-dev/database_data/Fields.csv') 
usethis::use_data(database_templates)
```

