---
title: "Preparing to use BarnebyLives!"
output: rmarkdown::html_vignette
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Preparing to use BarnebyLives!}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
description: >
  How to download and configure taxonomic and geographic datasets required
  for a working BarnebyLives instance.
---

# Introduction 

```{r, echo = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

```{r setup, message = F, echo = FALSE, eval = F}
# remotes::install_github('sagesteppe/BarnebyLives')
library(BarnebyLives)
library(tidyverse)
library(sf)
library(terra)
```

This vignette walks you through five steps:  
  - 1) Download taxonomic data (2-10 minutes, ~500MiB)  
  - 2) Download geographic data (a couple hours depending on extent)  
  - 3) Set up the instance (15-45 minutes processing) 
  - 4) Verify success

System requirements:
  - Stable internet connection for downloads
  - 4-20 GB available disk space (varies by geographic extent)
    - however the temporary objects, i.e. after downloading but before running
      `data_setup` will require significantly more memory. 
  - R version ≥ 4.0.0

You only need to do this once before using the package for analyses or label printing. 
Aside from the taxonomy module, none of this has to be re-ran. 

#  Decide where to store data

BarnebyLives relies on local copies of several large datasets to minimize API calls and ensure consistent performance. Storage requirements vary by geographic scope:

  - Regional collections (single state/province): 4-8 GB
  - Multi-state collections (e.g., Great Plains): 8-12 GB
  - Continental collections (e.g., Michigan to California): 16+ GB

The data persistence approach offers several advantages:

  - Reduced dependency on internet connectivity during analyses
  - Consistent data versions across analyses
  - Faster processing compared to repeated API calls
  - Offline capability for field work

It's important that you decide where BarnebyLives will be installed on your computer. 
Below, I will be downloading and copying some of the data to an external drive mounted to my system, where I have a directory named 'BL_sandbox'. 

```{r}
# Example directory (adjust for your system)
bl_dir <- "/media/steppe/hdd/BL_sandbox"
```

Note: All file paths in the pipeline can be specified independently, allowing you to relocate data after setup without breaking functionality.
So don't worry if you start an install somewhere you don't want the data to end up. 

## Taxonomic Data

BarnebyLives uses the *World Checklist of Vascular Plants* as its primary taxonomic backbone. 
This continuously updated resource from Kew Gardens provides standardized nomenclature for global plant diversity and serves as the foundation for *Plants of the World Online (POWO)*.

The `TaxUnpack` function downloads and processes WCVP data, creating three lookup tables:

Complete species catalog
Infraspecific taxa (varieties and subspecies)
Genus-level taxonomy

```{r read in WCVP data, eval = F}
## download and process the taxonomic data
p2 <- file.path(bl_dir, 'taxdata')

wcvp_update(p2) 
TaxUnpack(path = p2, bound = bound)
```

### IPNI 

Standard botanical author abbreviations from the International Plant Names Index (IPNI) ensure consistent citation formatting on specimen labels.

```{r Prepare material for authorship lookup tables, eval = F}
data('ipni_authors', package='BarnebyLives')
write.csv(ipni_authors, file.path(p2, 'ipni_author_abbreviations.csv'))
```

## Download all geographic data

### Automated downloads 

The `data_download` function retrieves most geographic data sets automatically. 

```{r data download geographic data, eval = F}
# I am using R markdown so want to be really sure save data to the correct location
# so I will use an absolute path
data_download(path = file.path(bl_dir, 'raw'))
# if you are using an interactive R script. (fileR) you should be able to do something
# similar or even just cd into the directory where you want to save everything and 
# run the function ala `download-data()`
```

## Manual downloads
Geographic data are organized in tiles following standard conventions. 
Define your study area as a bounding box using decimal degrees (WGS84/EPSG:4326 - default systems in most cartography software are close enough). 
Google Maps will work fine for this, just go a few miles outside of your extent to be sure. 

```{r determine tiles to download, eval = F}
bound <- data.frame(
   y = c( 48,  48,  41,  41,  48),
   x = c(-91, -82, -82, -91, -91)
 ) # note that the 5th pair of coordinates is the same
# as the first pair (48, -91), this is required to 'close'
# the square. 

tileSelector(bound)
```

`tile_selector` will help you determine which files you need to download. 
We will define a geographic extent that covers the area of interest. 
It will be a simple square, and grabbing coordinates from Open Street/Google/Apple Maps etc., or a GIS, will suffice.

### manual 

We will download a couple attributes from Geomorpho90m from a [Minio](https://opentopography.s3.sdsc.edu/minio/dataspace/OTDS.012020.4326.1/raster/) page. 
We need both '*aspect*' (NOT 'sine' or 'cosine') and '*geom*' from here, you should be able to control find ('ctrl+'f') these in their respective directories and manually download them.  
Once the files are downloaded copy them to the same location as you had `data_download` save files. 

Now we also need to manually download data on elevation from [MERIT-DEM](https://hydro.iis.u-tokyo.ac.jp/~yamadai/MERIT_DEM/).
Yes you technically need to register here, but you will automatically get a log in password to any email you enter, and it works!

These tiles follow the same naming convention as the previous data set - in fact Geomorpho90m is derived from MERIT-DEM one. 
Also save these files to the location you have `data_download` save files. 

## set up an instance for use 

You are now ready to set up an instance of BL for analysis. 

Specify where the downloaded data are `path`, where to save the date `pathOut`, the square (or bounding box) around your study area `bound`, and whether to remove the raw downloaded files 'cleanup'. 
I set `cleanup = FALSE`, until I am 100% certain that all downloads are good, which means running data through the pipeline and printing labels a couple times. 

```{r, eval = F}
data_setup( # this one will take maybe 30 minutes or so 
  path = file.path(bl_dir, 'raw'), 
  pathOut = file.path(bl_dir, 'geo'),
  bound = bound, cleanup = FALSE
  )
```

# Finishing up. 

You can check that all files are in place by running `` and you should be good to go forward
```{r}
# and you can check that everything has been set up using this. 
check_data_setup_outputs(file.path(bl_dir, 'geo'))
```

And that's it!
After a few runs of the system feel free to delete the raw data, you can do that by hand, or by initializing the instance using the `data_setup` function again. But that kind of seems like overkill!

```{r, eval = F}
data_setup( # this one will take maybe 30 minutes or so 
  path = file.path(bl_dir, 'raw'), 
  pathOut = file.path(bl_dir, 'geo'),
  bound = bound, cleanup = TRUE
  )
```

## Next Steps
With your BarnebyLives instance successfully configured, you can now:

- Validate taxonomic names against WCVP standards
- Extract environmental data for specimen localities

Consult the additional BarnebyLives vignettes for analysis workflows and advanced usage examples.
